---
title: "HW 5"
author: "Jake Vettoretti"
date: "2025-11-06"
output: github_document
---

```{r}
library(tidyverse)
set.seed(1)
```


## Problem 1


```{r}
# Define simulation function
bday_sim = function(n_room) {
  birthdays = sample(1:365, n_room, replace = TRUE)
  repeated_bday = length(unique(birthdays)) < n_room
  return(repeated_bday)
}

# Run simulation for group sizes 2–50, with 10,000 iterations each
bday_sim_results = 
  expand_grid(
    bdays = 2:50,
    iter = 1:10000
  ) |> 
  mutate(
    result = map_lgl(bdays, bday_sim) #returns logical vector
  ) |> 
  group_by(
    bdays
  ) |> 
  summarize(
    prob_repeat=mean(result)
  )

bday_sim_results

# Plot results
bday_sim_results |> 
  ggplot(aes(x = bdays, y = prob_repeat)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Group size (n)",
    y = "Probability of shared birthday",
    title = "Probability that at least two people share a birthday"
  ) +
  theme_minimal()
```

Comments:
The plot shows that the probability that at least two people share a birthday increases quickly as the number of people in the room grows. When only a few people are present the chance of a shared birthday is very small but the probability rises sharply as the group becomes larger. Around twenty three people the probability reaches about fifty percent and by forty people it is close to one hundred percent. This finding illustrates the birthday paradox where even small groups have a surprisingly high chance of sharing a birthday because the number of possible pairs increases rapidly with group size.


